A Decision Procedure for (Co)datatypes in SMT Solvers

### 2015-09-10 jour/jour.tex
starting to port

%%% TODO: In report, give small example where Single is needed

@@

\def\thewordpaper{paper}
\newcommand\dotReportFootnote[1]{.}

@@

\begin{rep}
\def\thewordpaper{report}
\renewcommand\dotReportFootnote[1]{.\footnote{#1}}
\end{rep}

@@

\begin{nomemoriam}
\noindent\hfill\emph{\phantom{In memoriam Morgan Deters 1979--2015}}

\vspace*{-6\smallskipamount}
\end{nomemoriam}
\begin{memoriam}
\noindent\hfill\emph{\relax{In memoriam Morgan Deters 1979--2015}}
\end{memoriam}

@@

\begin{nomemoriam}
\vskip3\smallskipamount
\end{nomemoriam}

@@

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground
constructor
terms, and only those. As a simple example, the coalgebraic specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
%(using an ML-like syntax)
{\looseness=-1
introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
in Peano notation but extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$.
The equation $\const{S}(\infty) \teq \infty$ holds as expected,
because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies
%the value~
$\infty$.
Compared with the conventional %algebraic
definition
\begin{conf}$\vthinspace \keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}$\vvthinspace, \end{conf}%
\begin{rep}\[\vthinspace \keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}\]\end{rep}%
%with a dedicated constructor for representing infinity, the
the codatatype avoids
one case by unifying the infinite and finite nonzero cases.

{\looseness=-1
Datatypes and codatatypes are an integral part of many proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have emerged in a few automatic theorem provers as well. The SMT-LIB
%\cite{barrett-et-al-2010}
format, implemented by most SMT
%(satisfiability modulo theories)
solvers, has been extended with a syntax for datatypes.
In this \thewordpaper, we introduce a
unified \hfill decision \hfill procedure \hfill for \hfill universal \hfill problems \hfill involving \hfill datatypes \hfill and \hfill codatatypes%
\begin{conf}\goodbreak\noindent\end{conf}%  %%% TYPESETTING BREAK
\begin{rep} \end{rep}%
in combination (Section~\ref{sec:the-decision-procedure}).
The procedure is described abstractly as a %tableau-like
calculus and is composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
It generalizes the procedure by Barrett et al.\ \cite{barrett-et-al-2007},
which covers only datatypes.
%To our knowledge, our procedure is the first of its kind for the theory of
%codatatypes.
\begin{conf}%
Detailed proofs %of completeness
are included in a %technical
report
\cite{our-report}.
\end{conf}%

Datatypes and codatatypes share many properties, so it makes sense
to consider them together. There are, however, at least three important
differences.
%
First, \emph{codatatypes need not be well-founded.}
For example, the type
$\vthinspace\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\allowbreak\, \const{SCons}(\tau,\allowbreak\: \ty{stream}_{\,\tau})\vthinspace$
of infinite sequences or streams over an element type $\tau$ is
allowed\begin{conf}, even though it has no base case\end{conf}\begin{rep}; the
corresponding datatype would be rejected as non-well-founded
\cite{blanchette-et-al-2015-esop}\end{rep}.
%
Second, \emph{a uniqueness rule takes the place of the acyclicity rule of datatypes.}
Cyclic constraints such as
$x \teq \const{S}(x)$ %, where $\const{C}$ is a constructor,
are unsatisfiable for datatypes,
thanks to an acyclicity rule,
but satisfiable for codatatypes. 
For the latter, a uniqueness rule ensures that two values
having the same infinite expansion must be equal; from $x
\teq \const{S}(x)$ and
$y \teq \const{S}(y)$, it deduces $x \teq y$.
These two rules are needed to ensure completeness
(solution soundness) on universal problems. %and the absence of spurious models.
They cannot be expressed as finite axiomatizations, so they naturally belong in
a decision procedure.
%    * in particular, acyclicity and uniqueness are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
%(For the other (co)datatype properties---the injectivity, distinctness, and
%exhaustiveness of constructors and the selector laws---it is
%widely recognized that decision procedures can be more efficient than
%axiomatizations.)
%axiom.
%
Third, \emph{it must be possible to express cyclic }(\emph{regular}) \emph{values as closed terms and
to enumerate them.} This is necessary both for finite model finding \cite{reynolds-et-al-2013}
%modulo theories
and for theory combinations. The $\mu$-binder notation associates a name with
a subterm; it is used to represent cyclic values in the generated
models\begin{rep} and in the metatheory\end{rep}. For example,
the $\mu$-term $\const{SCons}(1,\: \MU s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
stands for the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9,
\ldots$\afterLdots.

@@

\begin{rep}\paragraph{Related Work.} Barrett\end{rep}%
\begin{conf}Barrett\end{conf}
et al.\ \cite{barrett-et-al-2007} provide a good account of
related work on datatypes as of 2007, in addition to describing their
implementation in CVC3.
Since then, datatypes have been added not only to CVC4 (a complete rewrite of CVC3)
but also to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} \begin{rep}in unpublished work by Leonardo de Moura \end{rep}and \begin{rep}to \end{rep}a SPASS-like %prototypical
superposition prover \begin{rep}by Daniel Wand\end{rep} \cite{wand-2014}.
Closely related are the automatic structural
% cruanes-201x,wand-weidenbach-201x,  -- both were rejected
induction in both kinds of provers \cite{kersani-peltier-2013,reynolds-kuncak-2015},
the (co)datatype and (co)induction support in Dafny \cite{leino-moskal-2014},
and the (semi-)decision procedures for datatypes
implemented in Leon \cite{suter-et-al-2011} and RADA \cite{pham-et-al-2013}.
%
Datatypes are supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also generate models involving cyclic codatatype values.
Cyclic values have been studied extensively under the heading of
regular or rational trees---see Carayol and Morvan
\cite{carayol-morvan-2006} and Djellou et al.\ \cite{djelloul-et-al-2008}
for recent work.
The $\mu$-notation is inspired by the $\mu$-calculus
\begin{conf}\cite{kozen-1983}.\end{conf}
\begin{rep}\cite[etc.]{kozen-1983,endrullis-et-al-2011}.\end{rep}

@@

\paragraph{Conventions.}
Our setting is a monomorphic (or many-sorted) first-order logic.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are atomic sorts and interpreted as
nonempty domains. The set~$\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
\begin{rep}
The metavariables $\delta,\:\varepsilon$ range over (co)datatypes,
whereas $\tau,\:\upsilon$ range over arbitrary types.

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$).
\end{rep}
Names starting with an uppercase letter \begin{rep}(e.g.,
$\const{S}$) \end{rep}are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (for $n \ge 0$)
and a return type $\tau$, written
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$\begin{rep},
which collapses to $\const{f} : \tau$ if $n = 0$\end{rep}.
\begin{rep}%
Functions invocations $\const{f}(t_1,\ldots,t_n)$
apply the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols, called constants, can appear without
parentheses in terms.
\end{rep}
The set $\Funcs$ must at least contain
$\const{true},\, \const{false} : \ty{bool}$, interpreted as truth values.
The only predicate is equality ($\teq$)\begin{rep} and belongs to the logical symbols;
other predicates can be represented as functions to $\ty{bool}$,
with $\const{p}(\ldots)$ abbreviating $\const{p}(\ldots) \teq \const{true}$\end{rep}.
The notation $t\typ{\tau}$ stands for a term $t$ of type~$\tau$.
When applied to terms, the symbol $=$ denotes syntactic equality.
\begin{rep}The operator $\bigwedge_{\,i}\, \varphi_i$ abbreviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$.
Finally, $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$.\end{rep}

@@

%\paragraph{Specifications.}
In an SMT problem, the signature is typically given by specifying first the
uninterpreted %ordinary
types in any order, then the (co)datatypes with their constructors
and selectors in groups of mutually (co)recursive groups of (co)datatypes, and
finally any other function symbols.
%
Each (co)datatype specification consists of $\ell$~mutually recursive types that are
either all datatypes or all codatatypes. \begin{rep}Polymorphic types, nested
(co)recursion,\end{rep}\begin{conf}Nested (co)recursion\end{conf} and datatype--codatatype mixtures
fall outside this fragment%
\dotReportFootnote{%
In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
\begin{rep}We allow ourselves some notational parameterization
% at the metalevel
through subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}.\end{rep}

@@

Each (co)datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
\begin{conf}fresh\end{conf}\begin{rep}distinct and different from
existing names\end{rep}%
\dotReportFootnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by CVC4 and Z3, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(%\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}}
  \ldots)} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,n_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.
%
For types with several constructors, it is useful to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. Instead of extending $\Funcs$,
we let $\const{d}_{i\negvthinspace j}(t)$
be an abbreviation for
$t \teq \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(t), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(t)\bigr)$.
%This will simplify the presentation. % of the decision procedure.

\begin{rep}%
Here are a few examples of legal specifications of (co)datatype families:
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{llist}_{\,\tau} & \,=\, \const{LNil} \,\mid\, \const{LCons}(%\const{lhead}{:}\;
      \tau,\: %\const{ltail}{:}\;
      \ty{llist}_{\,\tau}) \\%[-.5\jot]
      \keyw{datatype}\;\, \ty{tree}_{\,\tau} & \,=\, \const{Node}(\tau,\:\, \ty{forest}_{\,\tau}) \\[-\jot]
      \keyw{and}\;\,\, \ty{forest}_{\,\tau} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree}_{\,\tau},\:\, \ty{forest}_{\,\tau})
\end{aligned}
\]

Because all types must be inhabited (nonempty), a datatype specification is
admissible only if a ground constructor term can be exhibited.
This rules out non-well-founded specifications such as
\[\keyw{datatype}\;\, \ty{fstream}_{\,\tau} \,=\, \const{FSCons}(\tau,\:\, \ty{fstream}_{\,\tau})\]
For codatatypes, no admissibility check is necessary because there is always a term,
finite or infinite, that witnesses nonemptiness \cite{blanchette-et-al-2015-esop}.
\end{rep}

A type $\delta$ depends on another type $\varepsilon$ if $\varepsilon$ is the
type of an argument to one of $\delta$'s constructors. Semantically, a set of
types is \emph{mutually} (\emph{co})\emph{recursive} if and only if the
associated dependency graph is strongly connected.
A type is (\emph{co})\emph{recursive} if it belongs to such a set of types.
\begin{rep}%
Types can be declared together in a mutual fashion even if they are not actually
(co)recursive. The semantic notion is more precise and is the one that interests
us.
\end{rep}%
%
Non(co)recursive type specifications such~as
\begin{conf}%
\vthinspace $\keyw{datatype}~\vthinspace\ty{option}_{\,\tau} \,=\, \const{None} \,\mid\, \const{Some}(\tau)$
\end{conf}%
\begin{rep}%
\[\begin{aligned}[t]
      \keyw{datatype}~\vthinspace\ty{option}_{\,\tau\!} & \,=\, \const{None} \,\mid\, \const{Some}(\tau) \\[-.5\jot]
      \keyw{codatatype}~\vthinspace\ty{complex} & \,=\, \const{Complex}(\const{re}{:}\; \ty{real},\; \const{im}{:}\;\ty{real})
\end{aligned}
\]
\end{rep}%
are permitted.
\begin{rep}At the semantic level, it makes no difference whether
such types are introduced as datatypes or as codatatypes.
\end{rep}%
%Without loss of generality, we consider that these types are datatypes
%instead of codatatypes.

@@

One way to characterize datatypes %semantically
is as the initial model of the
selector--constructor equations \cite{barrett-et-al-2007}.
\begin{rep}
A drawback of this approach is that it does
not naturally account for selectors applied to wrong constructors. Barrett et
al.\ address this by parameterizing the construction by default values, but
this gives rise to spurious equalities between unrelated terms---e.g.,
$\const{s}_1^1(\const{C}_2) \teq \const{s}_1^1(\const{C}_3)$. This flaw
could be corrected, but the added complexity seems to suggest that selectors
are better characterized axiomatically.

\end{rep}
A related semantic view of datatypes is as initial algebras. Codatatypes are
then defined dually as final coalgebras \cite{rutten-2000}. The datatypes are
generated by their constructors, whereas the codatatypes are viewed through
their selectors.
%By uniformly focusing on the constructors, the
%axiomatic approach emphasizes the commonality between datatypes and
%codatatypes, while sacrificing a theoretically fruitful notion of duality.

>>>>

%%% TODO: Give small example where Single is needed

\documentclass[smallcondensed,draft]{svjour3}

@@

@@

@@

\noindent\hfill\emph{\relax{In memoriam Morgan Deters 1979--2015}}

@@

@@

Codatatypes are freely generated by their constructors, but in contrast with datatypes,
infinit\-e constructor terms are also legitimate values for codatatypes
(Section~\ref{sec:co-datatypes}). Intuitively, the
values of a codatatype consist of all well-typed finite and infinite ground
constructor
terms, and only those. As a simple example, the coalgebraic specification
%
\[\keyw{codatatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat})\]
%
%(using an ML-like syntax)
introduces a type that
models the natural numbers $\const{Z}$, $\const{S}(\const{Z})$, $\const{S}(\const{S}(\const{Z}))$, $\ldots$\afterLdots{},
in Peano notation but extended with an
infinite value $\infty = \const{S}(\const{S}(\const{S}(\ldots)))$.
The equation $\const{S}(\infty) \teq \infty$ holds as expected,
because both sides expand to the infinite term
$\const{S}(\const{S}(\const{S}(\ldots)))$, which uniquely identifies
%the value~
$\infty$.
Compared with the conventional %algebraic
definition
\[\vthinspace \keyw{datatype}~\,\ty{enat} \,=\, \const{Z} \,\mid\, \const{S}(\ty{enat}) \,\mid\, \const{Infty}\]
%with a dedicated constructor for representing infinity, the
the codatatype avoids
one case by unifying the infinite and finite nonzero cases.

Datatypes and codatatypes are an integral part of many proof assistants,
including Agda, Coq, Isabelle, Matita, and PVS. In recent years, datatypes
have emerged in a few automatic theorem provers as well. The SMT-LIB
%\cite{barrett-et-al-2010}
format, implemented by most SMT
%(satisfiability modulo theories)
solvers, has been extended with a syntax for datatypes.
In this article, we introduce a
unified decision procedure for universal problems involving datatypes and codatatypes
in combination (Section~\ref{sec:the-decision-procedure}).
The procedure is described abstractly as a %tableau-like
calculus and is composable via the Nelson--Oppen method \cite{nelson-oppen-1979}.
It generalizes the procedure by Barrett et al.\ \cite{barrett-et-al-2007},
which covers only datatypes.
%To our knowledge, our procedure is the first of its kind for the theory of
%codatatypes.

Datatypes and codatatypes share many properties, so it makes sense
to consider them together. There are, however, at least three important
differences.
%
First, \emph{codatatypes need not be well-founded.}
For example, the type
$\vthinspace\keyw{codatatype}~\;\ty{stream}_{\,\tau} \,=\allowbreak\, \const{SCons}(\tau,\allowbreak\: \ty{stream}_{\,\tau})\vthinspace$
of infinite sequences or streams over an element type $\tau$ is
allowed, even though it has no base case; the
corresponding datatype would be rejected as non-well-founded
\cite{blanchette-et-al-2015-esop}.
%
Second, \emph{a uniqueness rule takes the place of the acyclicity rule of datatypes.}
Cyclic constraints such as
$x \teq \const{S}(x)$ %, where $\const{C}$ is a constructor,
are unsatisfiable for datatypes,
thanks to an acyclicity rule,
but satisfiable for codatatypes. 
For the latter, a uniqueness rule ensures that two values
having the same infinite expansion must be equal; from $x
\teq \const{S}(x)$ and
$y \teq \const{S}(y)$, it deduces $x \teq y$.
These two rules are needed to ensure completeness
(solution soundness) on universal problems. %and the absence of spurious models.
They cannot be expressed as finite axiomatizations, so they naturally belong in
a decision procedure.
%    * in particular, acyclicity and uniqueness are necessary for some proofs
%%      (and cannot be axiomatized finitely)
%    * and for model finding, without them we quickly get spurious models
%    * explain how finite model finding works
%(For the other (co)datatype properties---the injectivity, distinctness, and
%exhaustiveness of constructors and the selector laws---it is
%widely recognized that decision procedures can be more efficient than
%axiomatizations.)
%axiom.
%
Third, \emph{it must be possible to express cyclic }(\emph{regular}) \emph{values as closed terms and
to enumerate them.} This is necessary both for finite model finding \cite{reynolds-et-al-2013}
%modulo theories
and for theory combinations. The $\mu$-binder notation associates a name with
a subterm; it is used to represent cyclic values in the generated
models and in the metatheory. For example,
the $\mu$-term $\const{SCons}(1,\: \MU s.\; \const{SCons}(0,\: \const{SCons}(9,\: s)))$
stands for the lasso-shaped sequence $1, 0, 9, 0, 9, 0, 9,
\ldots$\afterLdots.

@@

\paragraph{Related Work.} Barrett
et al.\ \cite{barrett-et-al-2007} provide a good account of
related work on datatypes as of 2007, in addition to describing their
implementation in CVC3.
Since then, datatypes have been added not only to CVC4 (a complete rewrite of CVC3)
but also to the SMT solver
Z3 \cite{de-moura-bjoerner-2008} in unpublished work by Leonardo de Moura and to a SPASS-like %prototypical
superposition prover by Daniel Wand \cite{wand-2014}.
Closely related are the automatic structural
% cruanes-201x,wand-weidenbach-201x,  -- both were rejected
induction in both kinds of provers \cite{kersani-peltier-2013,reynolds-kuncak-2015},
the (co)datatype and (co)induction support in Dafny \cite{leino-moskal-2014},
and the (semi-)decision procedures for datatypes
implemented in Leon \cite{suter-et-al-2011} and RADA \cite{pham-et-al-2013}.
%
Datatypes are supported by the higher-order model finder Refute
\cite{weber-2008}. Its successor, Nitpick \cite{blanchette-nipkow-2010}, can
also generate models involving cyclic codatatype values.
Cyclic values have been studied extensively under the heading of
regular or rational trees---see Carayol and Morvan
\cite{carayol-morvan-2006} and Djellou et al.\ \cite{djelloul-et-al-2008}
for recent work.
The $\mu$-notation is inspired by the $\mu$-calculus
\cite[etc.]{kozen-1983,endrullis-et-al-2011}.

@@

\paragraph{Conventions.}
Our setting is a monomorphic (or many-sorted) first-order logic.
A signature $\Sig = (\Types, \Funcs)$ consists of a set of types $\Types$ and a
set of function symbols $\Funcs$. Types are atomic sorts and interpreted as
nonempty domains. The set~$\Types$ must contain a
distinguished type \ty{bool} interpreted as the set of truth
values. %, and may contain other interpreted types (e.g., \ty{int}, \ty{real}).
The metavariables $\delta,\:\varepsilon$ range over (co)datatypes,
whereas $\tau,\:\upsilon$ range over arbitrary types.

Function symbols are written in a sans-serif font (e.g., $\const{f}$, $\const{g}$) to
distinguish them from variables (e.g., $x$, $y$).
Names starting with an uppercase letter (e.g.,
$\const{S}$) are reserved for constructors. With each function symbol \const{f}
is associated a list of argument types $\tau_1,\ldots,\tau_n$ (for $n \ge 0$)
and a return type $\tau$, written
$\const{f} : \tau_1 \times \cdots \times \tau_n \to \tau$,
which collapses to $\const{f} : \tau$ if $n = 0$.
%
Functions invocations $\const{f}(t_1,\ldots,t_n)$
apply the $n$-ary function symbol
\const{f} to $n$ well-typed arguments $t_1 :\nobreak \tau_1$, \ldots, $t_n :
\tau_n$. Nullary function symbols, called constants, can appear without
parentheses in terms.
The set $\Funcs$ must at least contain
$\const{true},\, \const{false} : \ty{bool}$, interpreted as truth values.
The only predicate is equality ($\teq$) and belongs to the logical symbols;
other predicates can be represented as functions to $\ty{bool}$,
with $\const{p}(\ldots)$ abbreviating $\const{p}(\ldots) \teq \const{true}$.
The notation $t\typ{\tau}$ stands for a term $t$ of type~$\tau$.
When applied to terms, the symbol $=$ denotes syntactic equality.
The operator $\bigwedge_{\,i}\, \varphi_i$ abbreviates a conjunction
$\varphi_1 \mathrel\land \cdots \mathrel\land \varphi_n$.
Finally, $\bar x$ abbreviates a list or tuple $x_1,\ldots,x_n$.

@@

%\paragraph{Specifications.}
In an SMT problem, the signature is typically given by specifying first the
uninterpreted %ordinary
types in any order, then the (co)datatypes with their constructors
and selectors in groups of mutually (co)recursive groups of (co)datatypes, and
finally any other function symbols.
%
Each (co)datatype specification consists of $\ell$~mutually recursive types that are
either all datatypes or all codatatypes. Polymorphic types, nested
(co)recursion, and datatype--codatatype mixtures
fall outside this fragment.\footnote{In principle, rank-1 (top-level) polymorphism \cite{blanchette-paskevich-2013}
should not raise any special difficulties. Nesting datatypes inside datatypes,
and likewise for codatatypes, can be reduced to the mutual case
\cite{gunter-1993-not}. So the only genuinely interesting cases missing are
mixed nested (co)recursion as well as
(co)recursion through a non-(co)datatype (both of which make sense
\cite{blanchette-et-al-2014-codata}).}
We allow ourselves some notational parameterization
% at the metalevel
through subscripts---for example, $\ty{stream}_{\,\tau}$ denotes a
family of ground types including
$\ty{stream}_{\,\ty{int}}$, $\ty{stream}_{\,\ty{bool}}$,
and \smash{$\ty{stream}_{\,\ty{stream}_{\,\ty{real}}}$}.

@@

Each (co)datatype $\delta$ is equipped with
$m \ge 1$ constructors, and each constructor for $\delta$ takes zero or more
arguments and returns a $\delta$ value. The argument types must be either
ordinary, among the already known (co)datatypes, or among the (co)datatypes
being introduced.
%
To every argument corresponds a selector. The names for the (co)data\-types, the
constructors, and the selectors must be
distinct and different from existing names.%
\footnote{It can be convenient to specify the same selector
for several constructors associated with the same (co)data\-type,
as long as the argument types coincide. % \cite[Section~3]{blanchette-et-al-2014-codata}.
However, this is disallowed by CVC4 and Z3, so we do not consider it here.}
Schematically:
%
\[
\begin{aligned}[t]
\!(\keyw{co})\keyw{datatype}\;\,
  \delta_1 & {}=\, \smash{\const{C}_{11\!}(\bigl[\const{s}_{11\!}^1{:}\bigr]\vthinspace \tau_{11\!}^1, \ldots, \bigl[\const{s}_{11\!}^{n_{11\!}}{:}\bigr]\vthinspace \tau_{11\!}^{n_{11\!}})} \mid \cdots \mid \smash{\const{C}_{1m_1\!}(\ldots)} \\[-.5\jot]
   \smash{\vdots\,\,\,} \\[-1.5\jot]
  \keyw{and}\; \,\delta_{\elllx} & =\, \smash{\const{C}_{\elll 1\!}(%\bigl[\const{s}_{\elll 1\!}^1{:}\bigr]\vthinspace \tau_{\elll 1\!}^1, \ldots, \bigl[\const{s}_{\elll 1\!}^{n_{\elll 1\!}}{:}\bigr]\vthinspace \tau_{\elll 1\!}^{n_{\elll 1\!}}
  \ldots)} \mid \cdots \mid \smash{\const{C}_{\elll m_{\elllx}\!}(\ldots)}
\end{aligned}
\]
%
with
$\smash{\const{C}_{i\negvthinspace j} : \tau_{i\negvthinspace j}^1\times\cdots\times\tau_{i\negvthinspace j}^{\,n_{\smash{i\negvthinspace j}}} \to \delta_i}$
and $\smash{\const{s}_{i\negvthinspace j}^{\,k} : \delta_i \to \tau_{i\negvthinspace j}^{\,k}}$.
Defaults are assumed for the selector names if they are omitted.
The $\delta$ constructors and selectors are denoted by $\Ctr^\delta$ and
$\Sel^\delta$.
%
For types with several constructors, it is useful to provide discriminators
$\const{d}_{i\negvthinspace j} : \delta_i \to \ty{bool}$. Instead of extending $\Funcs$,
we let $\const{d}_{i\negvthinspace j}(t)$
be an abbreviation for
$t \teq \const{C}_{i\negvthinspace j}\bigl(\const{s}_{i\negvthinspace j}^1(t), \ldots, \const{s}_{i\negvthinspace j}^{\,n_{\smash{{i\negvthinspace j}}}}(t)\bigr)$.
%This will simplify the presentation. % of the decision procedure.

Here are a few examples of legal specifications of (co)datatype families:
\[\begin{aligned}[t]
      \keyw{codatatype}\; \,\ty{llist}_{\,\tau} & \,=\, \const{LNil} \,\mid\, \const{LCons}(%\const{lhead}{:}\;
      \tau,\: %\const{ltail}{:}\;
      \ty{llist}_{\,\tau}) \\%[-.5\jot]
      \keyw{datatype}\;\, \ty{tree}_{\,\tau} & \,=\, \const{Node}(\tau,\:\, \ty{forest}_{\,\tau}) \\[-\jot]
      \keyw{and}\;\,\, \ty{forest}_{\,\tau} & \,=\, \const{FNil} \,\mid\, \const{FCons}(\ty{tree}_{\,\tau},\:\, \ty{forest}_{\,\tau})
\end{aligned}
\]

Because all types must be inhabited (nonempty), a datatype specification is
admissible only if a ground constructor term can be exhibited.
This rules out non-well-founded specifications such as
\[\keyw{datatype}\;\, \ty{fstream}_{\,\tau} \,=\, \const{FSCons}(\tau,\:\, \ty{fstream}_{\,\tau})\]
For codatatypes, no admissibility check is necessary because there is always a term,
finite or infinite, that witnesses nonemptiness \cite{blanchette-et-al-2015-esop}.

A type $\delta$ depends on another type $\varepsilon$ if $\varepsilon$ is the
type of an argument to one of $\delta$'s constructors. Semantically, a set of
types is \emph{mutually} (\emph{co})\emph{recursive} if and only if the
associated dependency graph is strongly connected.
A type is (\emph{co})\emph{recursive} if it belongs to such a set of types.
Types can be declared together in a mutual fashion even if they are not actually
(co)recursive. The semantic notion is more precise and is the one that interests
us. Non(co)recursive type specifications such~as
\[\begin{aligned}[t]
      \keyw{datatype}~\vthinspace\ty{option}_{\,\tau\!} & \,=\, \const{None} \,\mid\, \const{Some}(\tau) \\[-.5\jot]
      \keyw{codatatype}~\vthinspace\ty{complex} & \,=\, \const{Complex}(\const{re}{:}\; \ty{real},\; \const{im}{:}\;\ty{real})
\end{aligned}
\]
are permitted.
At the semantic level, it makes no difference whether
such types are introduced as datatypes or as codatatypes.
%Without loss of generality, we consider that these types are datatypes
%instead of codatatypes.

@@

One way to characterize datatypes %semantically
is as the initial model of the
selector--constructor equations \cite{barrett-et-al-2007}.
A drawback of this approach is that it does
not naturally account for selectors applied to wrong constructors. Barrett et
al.\ address this by parameterizing the construction by default values, but
this gives rise to spurious equalities between unrelated terms---e.g.,
$\const{s}_1^1(\const{C}_2) \teq \const{s}_1^1(\const{C}_3)$. This flaw
could be corrected, but the added complexity seems to suggest that selectors
are better characterized axiomatically.

A related semantic view of datatypes is as initial algebras. Codatatypes are
then defined dually as final coalgebras \cite{rutten-2000}. The datatypes are
generated by their constructors, whereas the codatatypes are viewed through
their selectors.
%By uniformly focusing on the constructors, the
%axiomatic approach emphasizes the commonality between datatypes and
%codatatypes, while sacrificing a theoretically fruitful notion of duality.
